---
title: "Hands-on Exercise 08"
author: "Bhairavi Vairavelu"
date: "Oct 21 2024"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  message: false
  freeze: true
---

# Geographically Weighted Predictive Models

## 1.0 Overview

Predictive modelling uses statistical learning or machine learning techniques to predict outcomes. However, a set of known outcomes and predictors (variables) will be used to calibrate the predictive models.

In this exercise, we will build predictive models by using geographical random forest method

## 2.0 The Data

Aspatial data:

-   HDB Resale data

Geospatial data:

-   MP14_SUBZONE_WEB_PL

Locational factors with geographic coordinates:

-   Eldercare data

-   Hawker Centre data

-   Parks data

-   Supermarket data

-   CHAS clinics data

-   Childcare service data

-   Kindergartens data

-   MRT data

-   Bus stops data

Locational factors without geographic coordinates:

-   Primary schooldata

-   CBD coordinates

-   Shopping malls data

-   Good primary schools data

## 3.0 Installing & Loading R Packages

The code chunk below will check and install the R packages that we require for this exercise.

```{r}
pacman::p_load(sf, spdep, GWmodel, SpatialML, 
               tmap, rsample, Metrics, tidyverse)
```

## 4.0 Preparing the Data

### 4.1 Reading data file

We will read the input data set, which is a simple feature data frame.

```{r}
mdata <- read_rds("data/rds/mdata.rds")
```

### 4.2 Data Sampling

The entire data is split into training and test data sets, which are 65% and 35% respectively. This will be done using initial_split() of rsample package.

```{r}
set.seed(1234)
#resale_split <- rsample::initial_split(mdata, prop = 6.5/10,)
#train_data <- training(resale_split)
#test_data <- testing(resale_split)
#write_rds(train_data, "data/rds/train_data.rds")
#write_rds(test_data, "data/rds/test_data.rds")
```

x
