---
title: "Take-Home Exercise 03A"
author: "Bhairavi Vairavelu"
date: "Oct 21 2024"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  message: false
  freeze: true
---

# **Modelling Geography of Financial Inclusion with Geographically Weighted Methods**

## 1.0 Overview

Financial inclusion is the availability and equality of opportunities to access financial services. It refers to processes by which individuals and businesses can access appropriate, affordable, and timely financial products and services. These financial products and services include banking, loan, equity, and insurance products. It provides paths to enhance inclusiveness in economic growth by enabling the unbanked population to access the means for savings, investments, and insurance towards improving household income and reducing income inequality.

### 1.1 Tasks

In this exercise, we will be building an explanatory model to determine factors affecting financial inclusion by using geographical weighted regression methods.

### 1.2 Study Area

We will be focusing on the FinScope Tanzania 2023 dataset for this exercise, and conduct our analysis at the district level.

### 1.3 Analysis Method

Geographically Weighted Regression (GWR) is the spatial statistical technique that we will use in this exercise. GWR takes non-stationary variables into consideration, and models the local relationships between these independent variables, and an outcome of interest. For this exercise, the dependent variable is xx, and the independent variables are divided into either structural (?) and locational (?).

## 2.0 Data

For the purpose of this exercise, we will be using two datasets:

-   FinScope Tanzania 2023, which contains individual survey interview data

-   Tanzania geoBoundaries, which is a district level boundary GIS data

## 3.0 Installing and Launching Packages

We intend to utilize the following R packages in this exercise:

-   **olsrr**, for building OLS and performing diagnostic tests

-   **corrplot**, for multivariate data visualization and analysis

-   **ggpubr**, to customize ggplot2-based plots

-   **sf**, for spatial data handling

-   **spdep**, to create spatial weights matrix objects from polygon

-   **GWmodel**, for calibrating geographically weighted family of models

-   **tmap**, for visualizing geospatial objects

-   **tidyverse**, for attribute data handling,

-   **gtsummary**, to create analytical and summary tables

-   **ggstatsplot**, to create graphics with details from statistical tests included in the plots

-   **sfdep**, for computing spatial weights and autocorrelation statistics

The code chunk below can be used to install and launch the above-mentioned R packages:

```{r}
pacman::p_load(olsrr, corrplot, ggpubr, sf, spdep, GWmodel, tmap, tidyverse, gtsummary, ggstatsplot, sfdep)
```

## 4.0 Geospatial Data Wrangling

### 4.1 Importing Geospatial Data

The geospatial data used in this exercise, the Tanzania geoBoundaries district level data, is in ESRI shapefile format.

The code chunk below imports this shapefile by using st_read() of sf packages.

```{r}
gb = st_read(dsn = "data/geospatial", layer = "geoBoundaries-TZA-ADM2")
```

The report above shows that the R object used to contain the imported shapefile is called gb, and it is a simple feature object. The geometry type is multipolygon, and the gb simple feature object does not have any EPSG information. These multipolygon features are used to represent the geographic boundaries, and the GIS data is in WGS84 format. Also, the dataset contains 170 rows and 6 columns.

### 4.2 Updating CRS Information

It is noted that the correct EPSG code for Tanzania is EPSG: 4210, which can be verified here: LINK HERE

As such, we will use the code chunk below to update the newly imported gb simple feature data frame with the correct EPSG code.

```{r}
gbTransformed <- st_transform(gb, 4210)
```

After transforming the projection metadata, we can verify the projection of the newly transformed gbTransformed by using st_crs() of sf package. This can be achieved by using the code chunk below:

```{r}
st_crs(gbTransformed)
```

Note that the EPSG is indicated as 4210 now. We can now use the st_bbox() of sf package to reveal the extent of gbTransformed, as shown in the code chunk below:

```{r}
st_bbox(gbTransformed)
```

xx to elaborate what this st_bbox() does and what it means

To move this into EDA:

```{r}
tm_shape(gbTransformed) +
  tm_polygons()
```

x

## 5.0 Aspatial Data Wrangling

### 5.1 Importing Aspatial Data

The code chunk below uses read_csv() of readr package to import the csv file into R as a tibble data frame.

```{r}
tz = read_csv("data/aspatial/FinScope Tanzania 2023_Individual Main Data_FINAL.csv")
```

The report above shows that the R object used to contain the imported attribute data is called tz, and it is a tibble data frame. Also, the dataset contains 9915 rows and 721 columns.

After importing the file, it is important for us to examine if the file has been imported correctly. To do this, we will use glimpse() to display the data structure of tz.

```{r}
glimpse(tz)
```

We can also use summary() of base R to display the summary statistics of tz tibble data frame:

```{r}
summary(tz)
```

### 5.2 Data Cleanup

Given that the tz tibble data frame contains 721 columns, we first want to cleanup the dataset and only retain the fields that we want to focus our analysis on. Some of the fields we can consider for our analysis are:

1.  Overall Formal: this combines all variables that indicate formal financial service usage by an individual

2.  Informal: this combines all variables that indicate informal financial service usage by an individual

3.  Income Main: the main source of income for an individual

4.  Dedicated Farmer: whether the individual is a dedicated farmer or not, meaning more than 50% of their income originates from farming

5.  Banked: this combines all variables that indicate bank service usage by an individual

6.  \(MM\) Mobile: this combines all variables that indicate mobile money usage by an individual

7.  Insurance: this combines all variables that indicate any insurance usage by an individual

8.  (Form) Investments: this combines all variables that indicate formal investment mechanisms usage by an individual

9.  Pension: this combines all variables that indicate pension service usage by an individual

## 6.0 Exploratory Data Analysis

x

## 7.0 Financial Inclusion Model

x

## 8.0 Using gwModel

x
