---
title: "In-Class Exercise 05"
author: "Bhairavi Vairavelu"
date: "Sep 30 2024"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  message: false
  freeze: true
---

# Global and Local Measures of Spatial Autocorrelation

## 1. Getting Started

Four R packages will be used for this in-class exercise, they are: sf, sfdep, tmap and tidyverse.

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

## 2. The Data

Two datasets will be used

-   Hunan province administrative boundary layer at country level, a shapefile

-   Hunan's local development indicators in 2012, a csv file

### 2.1 Import shapefile into R

The code chunk below uses st_read() of sf package to import the Hunan shapefile into R.

```{r}
#|eval: false
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
write_rds(hunan, "data/rds/hunan.rds")
```

This code chunk will read the saved rds into R environment:

```{r}
hunan <- read_rds("data/rds/hunan.rds")
```

### 2.2 Import attribute table into R

Next, we will import Hunan_2012.csv into R using read_csv() of readr package.

```{r}
#|eval: false
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
write_rds(hunan2012, "data/rds/hunan2012.rds")
```

This code chunk will read the saved rds into R environment:

```{r}
hunan2012 <- read_rds("data/rds/hunan2012.rds")
```

### 2.3 Performing Relational Join

The code chunk below will be used to update the attribute table of hunan's SpatialPolygonsDataFrame with the attribute fields of hunan2012 data frame. This is performed using the left_join() of dplyr package.

```{r}
#|eval: false
hunan_joined <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
write_rds(hunan_joined, "data/rds/hunan_joined.rds")
```

This code chunk will read the saved rds into R environment:

```{r}
hunan_joined <- read_rds("data/rds/hunan_joined.rds")
```

### 2.4 Plotting a choropleth map

We will plot a choropleth map showing the distribution of GDPPC of Hunan province.

```{r}
tmap_mode("plot")
tm_shape(hunan_joined) +
  tm_fill("GDPPC", 
          style = "quantile", 
          palette = "Blues",
          title = "GDPPC") +
  tm_layout(main.title = "Distribution of GDP per capita by county, Hunan Province",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2)
```

## 3. Global Measures of Spatial Association

### 3.1 Deriving Queen's contiguity weights

```{r}
wm_q <- hunan_joined %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb,
                         style = "W"),
         .before = 1) 
wm_q
```

### 3.2 Computing Global Moran I

In the code chunk below, global_moran() function is used to compute the Moran’s I value. Different from spdep package, the output is a tibble data.frame.

```{r}
moranI <- global_moran(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)
glimpse(moranI)
```

### 3.3 Performing Global Moran's I test

```{r}
global_moran_test(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)
```

### 3.4 Performing Global Moran's I permutation test

```{r}
set.seed(1234)
global_moran_perm(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt,
                  nsim = 99)
```

### 3.5 Computing Local Moran's I

```{r}
lisa <- wm_q %>% 
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_moran)
```

Visualizing local Moran's I:

```{r}
tmap_mode("plot")
tm_shape(lisa) +
  tm_fill("ii") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8)) +
  tm_layout(
    main.title = "local Moran's I of GDPPC",
    main.title.size = 2)
```

Visualizing p-value of local Moran's I:

```{r}
tmap_mode("plot")
tm_shape(lisa) +
  tm_fill("p_ii_sim") + 
  tm_borders(alpha = 0.5) +
   tm_layout(main.title = "p-value of local Moran's I",
            main.title.size = 2)
```

To plot both of them together:

```{r}
tmap_mode("plot")
map1 <- tm_shape(lisa) +
  tm_fill("ii") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8)) +
  tm_layout(main.title = "local Moran's I of GDPPC",
            main.title.size = 0.8)

map2 <- tm_shape(lisa) +
  tm_fill("p_ii",
          breaks = c(0, 0.001, 0.01, 0.05, 1),
              labels = c("0.001", "0.01", "0.05", "Not sig")) + 
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "p-value of local Moran's I",
            main.title.size = 0.8)

tmap_arrange(map1, map2, ncol = 2)
```

Plotting LISA map:

```{r}
lisa_sig <- lisa  %>%
  filter(p_ii_sim < 0.05)
tmap_mode("plot")
tm_shape(lisa) +
  tm_polygons() +
  tm_borders(alpha = 0.5) +
tm_shape(lisa_sig) +
  tm_fill("mean") + 
  tm_borders(alpha = 0.4)
```

### 3.6 Hot & Cold Spot Area Analysis

To derive a spatial weight matrix:

```{r}
wm_idw <- hunan_joined %>%
  mutate(nb = include_self(
    st_contiguity(geometry)),
    wts = st_inverse_distance(nb, 
                              geometry, 
                              scale = 1,
                              alpha = 1),
         .before = 1)
```

Computing local Gi statistics:

```{r}
HCSA <- wm_idw %>% 
  mutate(local_Gi = local_gstar_perm(
    GDPPC, nb, wts, nsim = 99),
         .before = 1) %>%
  unnest(local_Gi)
HCSA
```

Visualizing Gi:

```{r}
tmap_mode("plot")
tm_shape(HCSA) +
  tm_fill("gi_star") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8))
```

Visualizing p-value of HCSA:

```{r}
tmap_mode("plot")
tm_shape(HCSA) +
  tm_fill("p_sim") + 
  tm_borders(alpha = 0.5)
```

Visualizing local HCSA:

```{r}
tmap_mode("plot")
map1 <- tm_shape(HCSA) +
  tm_fill("gi_star") + 
  tm_borders(alpha = 0.5) +
  tm_view(set.zoom.limits = c(6,8)) +
  tm_layout(main.title = "Gi* of GDPPC",
            main.title.size = 0.8)

map2 <- tm_shape(HCSA) +
  tm_fill("p_value",
          breaks = c(0, 0.001, 0.01, 0.05, 1),
              labels = c("0.001", "0.01", "0.05", "Not sig")) + 
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "p-value of Gi*",
            main.title.size = 0.8)

tmap_arrange(map1, map2, ncol = 2)
```

Visualizing hot spot and cold spot areas:

```{r}
HCSA_sig <- HCSA  %>%
  filter(p_sim < 0.05)
tmap_mode("plot")
tm_shape(HCSA) +
  tm_polygons() +
  tm_borders(alpha = 0.5) +
tm_shape(HCSA_sig) +
  tm_fill("cluster") + 
  tm_borders(alpha = 0.4)
```
