---
title: "In-Class Exercise 04"
author: "Bhairavi Vairavelu"
date: "Sep 16 2024"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  message: false
  freeze: true
---

# Geographically Weighted Summary Statistics

## 1. Getting Started

Let's ensure that the following packages of R are installed and launched in R.

```{r}
pacman::p_load(sf, ggstatsplot, tmap, tidyverse, knitr, GWmodel)
```

## 2. Loading the Data

The code chunk below uses st_read() of sf package to import Hunan shapefile into R. This imported shapefile will be a simple feature object of sf.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

Next, we will import the Hunan_2012.csv into R by using (read_csv) of readr package. The output is a R dataframe class.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

## 3. Preparing the Data

The code chunk below will be used to update the attribute table of hunan's SpatialPolygonsDataFrame with the attribute fields of hunan2012 dataframe. This will be performed using the left_join() of dplyr package.

```{r}
hunan_sf <- left_join(hunan,hunan2012)%>%
  select(1:3, 7, 15:16, 31:32)
```

## 4. Converting to SpatialPolygonDataFrame

GWmodel package can only handle spatial data in older sp format, and not the sf format.

```{r}
hunan_sp <- hunan_sf %>%
  as_Spatial()
```

## 5. Geographically Weighted Summary Statistics with Adaptive Bandwidth

Cross Validation:

```{r}
bw_CV <- bw.gwr(GDPPC ~ 1,
                data = hunan_sp,
                approach = "CV",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T)
```

AIC:

```{r}
bw_AIC <- bw.gwr(GDPPC ~ 1,
                data = hunan_sp,
                approach = "AIC",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T)
```

Bandwidth AIC:

```{r}
gwstat <- gwss(data = hunan_sp,
               vars = "GDPPC",
               bw = bw_AIC,
               kernel = "bisquare",
               adaptive = TRUE,
               longlat = T)
```

Preparing the output data:

```{r}
gwstat_df <- as.data.frame(gwstat$SDF)
hunan_gstat <- cbind(hunan_sf, gwstat_df)
```

Visualizing it:

```{r}
tm_shape(hunan_gstat) +
  tm_fill("GDPPC_LM",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Distribution of geographically wieghted mean",
            main.title.position = "center",
            main.title.size = 2.0,
            legend.text.size = 1.2,
            legend.height = 1.50, 
            legend.width = 1.50,
            frame = TRUE)
```

