---
title: "In-Class Exercise 04"
author: "Bhairavi Vairavelu"
date: "Sep 16 2024"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  message: false
  freeze: true
---

# Geographically Weighted Summary Statistics

## 1. Getting Started

Let's ensure that the following packages of R are installed and launched in R.

```{r}
pacman::p_load(sf, ggstatsplot, tmap, tidyverse, knitr, GWmodel)
```

## 2. Loading the Data

The code chunk below uses st_read() of sf package to import Hunan shapefile into R. This imported shapefile will be a simple feature object of sf.

```{r}
#|eval: false
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
write_rds(hunan, "data/rds/hunan.rds")
```

To read in the rds:

```{r}
hunan <- read_rds("data/rds/hunan.rds")
```

Next, we will import the Hunan_2012.csv into R by using (read_csv) of readr package. The output is a R dataframe class.

```{r}
#|eval: false
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
write_rds(hunan2012, "data/rds/hunan2012.rds")
```

To read in the rds:

```{r}
hunan2012 <- read_rds("data/rds/hunan2012.rds")
```

## 3. Preparing the Data

The code chunk below will be used to update the attribute table of hunan's SpatialPolygonsDataFrame with the attribute fields of hunan2012 dataframe. This will be performed using the left_join() of dplyr package.

```{r}
#|eval: false
hunan_sf <- left_join(hunan,hunan2012)%>%
  dplyr::select(1:3, 7, 15:16, 31:32)
write_rds(hunan_sf, "data/rds/hunan_sf.rds")
```

To read in the rds:

```{r}
hunan_sf <- read_rds("data/rds/hunan_sf.rds")
```

## 4. Converting to SpatialPolygonDataFrame

GWmodel package can only handle spatial data in older sp format, and not the sf format.

```{r}
#|eval: false
hunan_sp <- hunan_sf %>%
  as_Spatial()
write_rds(hunan_sp, "data/rds/hunan_sp.rds")
```

To read in the rds:

```{r}
hunan_sp <- read_rds("data/rds/hunan_sp.rds")
```

## 5. Geographically Weighted Summary Statistics with Adaptive Bandwidth

Cross Validation:

```{r}
bw_CV <- bw.gwr(GDPPC ~ 1,
                data = hunan_sp,
                approach = "CV",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T)
```

AIC:

```{r}
bw_AIC <- bw.gwr(GDPPC ~ 1,
                data = hunan_sp,
                approach = "AIC",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T)
```

Bandwidth AIC:

```{r}
gwstat <- gwss(data = hunan_sp,
               vars = "GDPPC",
               bw = bw_AIC,
               kernel = "bisquare",
               adaptive = TRUE,
               longlat = T)
```

Preparing the output data:

```{r}
#|eval: false
gwstat_df <- as.data.frame(gwstat$SDF)
hunan_gstat <- cbind(hunan_sf, gwstat_df)
write_rds(gwstat_df, "data/rds/gwstat_df.rds")
write_rds(hunan_gstat, "data/rds/hunan_gstat.rds")
```

To read in the rds:

```{r}
gwstat_df <- read_rds("data/rds/gwstat_df.rds")
hunan_gstat <- read_rds("data/rds/hunan_gstat.rds")
```

Visualizing it:

```{r}
tm_shape(hunan_gstat) +
  tm_fill("GDPPC_LM",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Distribution of geographically wieghted mean",
            main.title.position = "center",
            main.title.size = 2.0,
            legend.text.size = 1.2,
            legend.height = 1.50, 
            legend.width = 1.50,
            frame = TRUE)
```
