---
title: "In-Class Exercise 01"
author: "Bhairavi Vairavelu"
date: "Aug 26 2024"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  message: false
  freeze: true
---

## 1. Getting Started

### 1.1 Loading the R Packages

This code chunk loads the R packages we require for this exercise:
```{r}
pacman::p_load(tidyverse, sf, tmap, ggstatsplot)
```

### 1.2 Loading the Data

This code chunk imports the shapefile:
```{r}
mpsz14_shp <- st_read(dsn = "data", 
                     layer = "MP14_SUBZONE_WEB_PL")
```

This code chunk (should) import the kml file. However, since it's corrupted we will not be using it.
```{r}
# mpsz14_kml <- st_read("data/MasterPlan2014SubzoneBoundaryWebKML.kml")
```

We could help GovTech by writing out a new kml file and send it back to them :)
```{r}
#| output: false
st_write(mpsz14_shp,
         "data/MP14_SUBZONE_WEB_PL.kml",
         delete_dsn = TRUE)
```

This code chunk imports the shapefile (unable to find this in data.gov):
```{r}
#mpsz19_shp <- st_read(dsn = "data", 
#                     layer = "MP14_SUBZONE_WEB_PL")
```

This code chunk imports the kml file (unable to load this):
```{r}
#mpsz19_kml <- st_read("data/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

## 2. Reprojection

### 1,3 Transforming Coordinate System

This code chunk transform the coordinate system:
```{r}
mpsz19_shp <- st_read(dsn = "data", 
                     layer = "MP14_SUBZONE_WEB_PL") %>%
  st_transform(crs = 3414)
```

